---
title: "667Final"
output:
  html_document: default
  pdf_document: default
date: "2025-11-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Data Setup & Visualization

```{r}
library(readr)
library(tidyverse)
library(pROC)
library(geepack)   # for GEE
library(lme4)      # for GLMM (lmer, glmer)
library(broom)     # for tidy() tables

########################################
## 1. Read & Clean Data
########################################

oasis_raw <- read.csv("oasis_longitudinal.csv")
oasis <- oasis_raw %>%
  rename(
    subject_id = Subject.ID,
    visit      = Visit,
    mr_delay   = MR.Delay,
    sex        = M.F,
    age        = Age,
    educ       = EDUC,
    mmse       = MMSE,
    cdr        = CDR,
    group      = Group
  ) %>%
  mutate(

    # Outcome: Poor Cognition
    poor_cognition = if_else(mmse < 24, 1L, 0L),

    # Factor conversions
    sex = factor(sex, levels = c("F", "M")),
    group = factor(group, levels = c("Nondemented", "Demented")),
    subject_id = factor(subject_id),

    # -----------------------------
    # OPTION A: Education Categories
    # -----------------------------
    educ_cat = case_when(
      educ <= 12 ~ "Low",
      educ >= 13 & educ <= 15 ~ "Medium",
      educ >= 16 ~ "High",
      TRUE ~ NA_character_
    ),
    
    # Convert to ordered factor so "Low" is reference
    educ_cat = factor(educ_cat, 
                      levels = c("Low", "Medium", "High"))
  )



########################################
## 4. Build Dataset for GEE (all visits)
########################################

# Keep rows with all key vars present for GEE models
oasis_gee <- oasis %>%
  filter(
    !is.na(mmse),
    !is.na(age),
    !is.na(educ),
    !is.na(sex),
    !is.na(cdr),
    !is.na(group)
  ) %>%
  arrange(subject_id) %>%
  group_by(subject_id) %>%
  mutate(
    n_visits = n()   # just for info / optional sensitivity check
  ) %>%
  ungroup()


```

```{r}

library(ggplot2)

ggplot(oasis_gee, aes(x = visit, y = mmse, group = subject_id, color = educ_cat)) +

  geom_line(alpha = 0.3, size = 0.5) +
  
  geom_smooth(aes(group = educ_cat), method = "loess", se = FALSE, size = 1) +
  labs(
    title = "Visualization of Raw MMSE Trajectories",
    subtitle = "Spaghetti plot with group mean trends",
    x = "Visit",
    y = "MMSE Score"
  ) +
  theme_minimal()
```

# Methods



# 2. GEE Analysis for RQ1 (Binary Outcome: Poor Cognition)


## 2.1 Model Specification and Fitting

We modelled the binary indicator of poor cognition (MMSE \< 24) using generalized estimating equations (GEE) with a binomial variance and logit link function, specifying subject as the clustering unit. This models the log-odds of poor cognition as a linear function of visit, age, education, and sex.

$$\text{logit}(P(Y_{it}=1)) = \beta_0 + \beta_1 \text{Visit}_{it} + \beta_2 \text{Education_Level}_i + \beta_3 (\text{Visit}_{it} \times \text{Education_Level}_i) + \beta_4 \text{AgeBaseline}_i  + \beta_5 \text{Sex}_i$$

```{r}
library(geepack)
library(broom)

form_rq1 <- poor_cognition ~ visit * educ_cat + age + sex

gee_rq1_ind <- geeglm(
  formula = form_rq1,
  family  = binomial(link = "logit"),
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "independence"
)

gee_rq1_exch <- geeglm(
  formula = form_rq1,
  family  = binomial(link = "logit"),
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "exchangeable"
)

gee_rq1_ar1 <- geeglm(
  formula = form_rq1,
  family  = binomial(link = "logit"),
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "ar1"
)

summary(gee_rq1_ind)
summary(gee_rq1_exch)

summary(gee_rq1_ar1)

```

## 2.2 Diagnostics: Standard Errors and Empirical Correlation

```{r}
#| label: gee-rq1-extract-se

# Use the Exchangeable model for diagnostics
gee_exch <- gee_rq1_exch
tidy(gee_rq1_exch, conf.int = TRUE)
# 1. Extract Naive and Robust Standard Errors
se_naive <- sqrt(diag(gee_exch$geese$vbeta.naiv)) # Naive SE (based on model structure assumption)
se_robust <- sqrt(diag(gee_exch$geese$vbeta))    # Robust SE (based on empirical data variability)

# 2. Prepare for Small Sample Bias Correction (BC)
N <- length(unique(oasis_gee$subject_id)) # Number of clusters (subjects)
p <- length(coef(gee_exch))               # Number of model parameters
# BC Factor: sqrt(N / (N - p))
bc_factor <- sqrt(N / (N - p))

# 3. Apply Small Sample Bias Correction
se_robust_bc <- se_robust * bc_factor

# 4. Compile coefficient and SE comparison table
coef_est <- coef(gee_exch)
se_table <- data.frame(
  estimate = coef_est,
  SE_naive = se_naive,
  SE_robust = se_robust,
  SE_robust_bc = se_robust_bc
) %>%
  mutate(
    Z_score = estimate / SE_robust_bc,
    P_value_BC = 2 * (1 - pnorm(abs(Z_score))) # Calculate P-value using Z-score and BC SE
  )

# 5. Extract estimated working correlation (alpha)
alpha_exch <- gee_exch$geese$alpha

print(paste("Estimated Working Correlation (Exchangeable Alpha):", round(alpha_exch, 4)))

knitr::kable(
  se_table,
  digits = 3,
  caption = "GEE Coefficient and Standard Error Comparison (RQ1: poor_cognition)"
) %>%
  kable_styling()

```

```{r}
#| label: gee-rq1-empirical-corr

# 1. Extract Pearson residuals from the Exchangeable model
oasis_gee <- oasis_gee %>%
  arrange(subject_id, visit) %>%
  mutate(resid_exch = residuals(gee_rq1_exch, type = "pearson"))

# 2. Define Lag range (up to 5 visits apart)
lags <- 1:5 

# 3. Calculate empirical correlation for each lag
empirical_corr <- purrr::map_dfr(lags, function(h) {
  # Create lagged residual pairs within each subject
  tmp <- oasis_gee %>%
    group_by(subject_id) %>%
    mutate(resid_lead = lead(resid_exch, h)) 
    
  # Filter pairs where both residuals exist (not NA)
  tmp <- tmp %>%
    ungroup() %>%
    filter(!is.na(resid_lead)) 

  # Calculate correlation across all pairs
  data.frame(
    lag = h,
    N_pairs = nrow(tmp),
    corr = cor(tmp$resid_exch, tmp$resid_lead)
  )
})

knitr::kable(
  empirical_corr,
  digits = 3,
  caption = "Empirical Residual Correlation by Lag (RQ1)"
)
```

GLMM is better

```{r}

library(tibble)
library(kableExtra)
library(gt)
qic_tab <- tibble(
  corstr = c("independence", "exchangeable", "ar1"),
  QIC    = c(
    geepack::QIC(gee_rq1_ind)[["QIC"]],
    geepack::QIC(gee_rq1_exch)[["QIC"]],
    geepack::QIC(gee_rq1_ar1)[["QIC"]]
  )
)
qic_tab %>%
  gt() %>%
  tab_header(
    title = "Model Fit Comparison",
    subtitle = "QIC Values by Correlation Structure For RQ1"
  )
# kable(qic_tab, caption = "Table 1: QIC Comparison of Correlation Structures For RQ1")
tidy_gee_rq1_exch <- broom::tidy(gee_rq1_exch) %>% 
  mutate(
    OR   = exp(estimate),
    LCL  = exp(estimate - 1.96 * std.error),
    UCL  = exp(estimate + 1.96 * std.error)
  )

```

```{r}

summary(gee_rq1_exch)
```

```{r}
#| label: gee-rq1-coef-comp-fixed
library(dplyr)
library(broom)
library(purrr)
library(tidyr)

models <- list(
  Independence = gee_rq1_ind, 
  Exchangeable = gee_rq1_exch, 
  AR1 = gee_rq1_ar1
)

coef_comp <- map_dfr(models, tidy, .id = "model_name") %>%
  filter(term %in% c("(Intercept)", "visit", "educ_catHigh", "educ_catMedium", 
                     "visit:educ_catHigh", "visit:educ_catMedium")) %>%
  dplyr::select(model_name, term, estimate) %>%
  pivot_wider(names_from = model_name, values_from = estimate)

knitr::kable(
  coef_comp, 
  digits = 3, 
  caption = "Step 7: Robustness of Coefficient Estimates across Working Correlation Structures"
)
```

```{r}
#| label: gee-rq1-prediction-plot
library(ggplot2)

# 1. Final model Exchangeable
final_model <- gee_rq1_exch

# 2. Prediction Grid
new_data <- expand.grid(
  visit = oasis_gee$visit,
  age    = mean(oasis_gee$age, na.rm = TRUE),
  educ_cat = oasis_gee$educ_cat,
  sex    = factor("F", levels = levels(oasis_gee$sex)) 
)


preds <- predict(final_model, newdata = new_data, type = "response")

plot_data <- new_data %>%
  mutate(pred_prob = as.numeric(preds))

# 4. plot
ggplot(plot_data, aes(x = visit, y = pred_prob, color = educ_cat, linetype = educ_cat)) +
  geom_line(size = 1.2) +
  labs(
    title = "Population-Average Probability of Poor Cognition Over Visit",
    subtitle = "Based on GEE (Exchangeable Correlation)",
    y = "Probability of Poor Cognition (MMSE < 24)",
    x = "Visit Number",
    color = "Educ_level",
    linetype = "Educ_level"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) # Y轴设为 0-100%
```

# 3. GEE Analysis for RQ2 (Continuous Outcome: MMSE)

## 3.1 Model Specification and Fitting

$$E(MMSE) = \beta_0 + \beta_1 \text{Visit}_{it} + \beta_2 \text{Education_Level}_i + \beta_3 (\text{Visit}_{it} \times \text{Education_Level}_i) + \beta_4 \text{AgeBaseline}_i  + \beta_5 \text{Sex}_i$$


```{r}
library(geepack)
library(broom)
library(tibble)
library(dplyr)

# Mean model for RQ2
form_rq2 <- mmse ~ visit * educ_cat + age  + sex

gee_rq2_ind <- geeglm(
  formula = form_rq2,
  family  = gaussian,
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "independence"
)

gee_rq2_exch <- geeglm(
  formula = form_rq2,
  family  = gaussian,
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "exchangeable"
)

gee_rq2_ar1 <- geeglm(
  formula = form_rq2,
  family  = gaussian,
  data    = oasis_gee,
  id      = subject_id,
  corstr  = "ar1"
)
summary(gee_rq2_ind)
summary(gee_rq2_exch)   # for a first look
summary(gee_rq2_ar1)
```

## 3.2 Diagnostics and Model Selection (QIC)

```{r}
qic_rq2 <- tibble(
  corstr = c("independence", "exchangeable", "ar1"),
  QIC    = c(
    geepack::QIC(gee_rq2_ind)[["QIC"]],
    geepack::QIC(gee_rq2_exch)[["QIC"]],
    geepack::QIC(gee_rq2_ar1)[["QIC"]]
  )
)

tidy(gee_rq2_exch,conf.int = TRUE)
library(knitr)

qic_rq2 %>%
  gt() %>%
  tab_header(
    title = "Model Fit Comparison",
    subtitle = "QIC Values by Correlation Structure For RQ2"
  )
```

```{r}
gee_rq2_best <- gee_rq2_exch  # or _exch / _ind depending on QIC
summary(gee_rq2_best)
tidy_gee_rq2 <- broom::tidy(gee_rq2_best) %>%
  mutate(
    LCL = estimate - 1.96 * std.error,
    UCL = estimate + 1.96 * std.error
  )

tidy_gee_rq2

```

```{r}
#| label: gee-rq2-extract-se

gee_exch_mmse <- gee_rq2_exch

se_naive <- sqrt(diag(gee_exch_mmse$geese$vbeta.naiv))
se_robust <- sqrt(diag(gee_exch_mmse$geese$vbeta))

N <- length(unique(oasis_gee$subject_id)) # 150
p <- length(coef(gee_exch_mmse))
bc_factor <- sqrt(N / (N - p))
se_robust_bc <- se_robust * bc_factor

coef_est <- coef(gee_exch_mmse)
se_table_rq2 <- data.frame(
  estimate = coef_est,
  SE_naive = se_naive,
  SE_robust = se_robust,
  SE_robust_bc = se_robust_bc
) %>%
  mutate(
    Z_score = estimate / SE_robust_bc,
    P_value_BC = 2 * (1 - pnorm(abs(Z_score)))
  )

alpha_rq2 <- gee_exch_mmse$geese$alpha

print(paste("RQ2 Estimated Working Correlation (MMSE Alpha):", round(alpha_rq2, 4)))

knitr::kable(
  se_table_rq2,
  digits = 3,
  caption = "RQ2: GEE Coefficient and Standard Error Comparison (Outcome: MMSE)"
) %>%
  kable_styling()
```

```{r}
#| label: gee-rq2-empirical-corr
library(dplyr)
library(purrr)

oasis_gee <- oasis_gee %>%
  arrange(subject_id, visit) %>%
  mutate(resid_mmse = residuals(gee_rq2_exch, type = "pearson"))


lags <- 1:4


empirical_corr_rq2 <- purrr::map_dfr(lags, function(h) {
  tmp <- oasis_gee %>%
    group_by(subject_id) %>%
    mutate(resid_lead = lead(resid_mmse, h)) %>%
    ungroup() %>%
    filter(!is.na(resid_lead))

  data.frame(
    lag = h,
    N_pairs = nrow(tmp),
    corr = cor(tmp$resid_mmse, tmp$resid_lead)
  )
})

knitr::kable(
  empirical_corr_rq2,
  digits = 3,
  caption = "RQ2: Empirical Residual Correlation by Lag (MMSE)"
)
```

```{r}

estimated_alpha <- summary(gee_rq2_exch)$geese$alpha

ggplot(empirical_corr_rq2, aes(x = lag, y = corr)) +
  geom_point(size = 3) + 
  geom_line() +
  geom_hline(yintercept = estimated_alpha, color = "red", linetype = "dashed") +
  scale_y_continuous(limits = c(-0.2, 1)) +
  labs(
    title = "Empirical Correlogram vs Working Correlation",
    subtitle = "If Exchangeable is correct, points should hover around the red dashed line",
    x = "Lag (Visit Distance)",
    y = "Correlation of Pearson Residuals"
  ) +
  theme_minimal()
```

# Model diagnosis

```{r}
# --- Model diagnosis ---

## 5.1 Residuals vs Fitted

# 提取拟合值和残差
diag_data <- oasis_gee %>%
  mutate(
    fitted = fitted(gee_rq2_exch),
    resid  = residuals(gee_rq2_exch, type = "pearson")
  )

ggplot(diag_data, aes(x = fitted, y = resid, color = educ_cat)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, color = "black", size = 0.5) + # 添加平滑线辅助判断
  labs(
    title = "Pearson Residuals vs Fitted Values",
    x = "Fitted MMSE",
    y = "Pearson Residuals"
  ) +
  theme_minimal()

## 5.2 影响点诊断 (Leave-One-Cluster-Out Analysis)
# 警告: 计算量较大，需要循环拟合 N 次 (N=150)

# 定义一个函数，移除某个 Subject 后重新拟合模型并提取主要系数
check_influence <- function(drop_id, full_data) {
  # 排除特定 ID
  sub_dat <- filter(full_data, subject_id != drop_id)
  
  # 重新拟合 (使用 tryCatch 避免报错中断)
  tryCatch({
    mod <- geeglm(mmse ~ visit * educ_cat + age  + sex, 
                  family = gaussian, data = sub_dat, id = subject_id, 
                  corstr = "exchangeable")
    # 提取我们最关心的交互项系数: visit:groupDemented
    coef(mod)["visit:educ_catHigh"]
  }, error = function(e) NA)
}

# 获取所有 ID
all_ids <- unique(oasis_gee$subject_id)

# 运行循环 (使用 purrr::map_dbl 简化)
# 注意: 这可能需要几秒到一分钟
influence_res <- map_dbl(all_ids, check_influence, full_data = oasis_gee)

# 创建数据框并计算 Delta Beta (变化量)
full_coef <- coef(gee_rq2_exch)["visit:educ_catHigh"]
inf_df <- data.frame(
  subject_id = all_ids,
  coef_jack  = influence_res
) %>%
  mutate(delta_beta = coef_jack - full_coef)

# 绘制影响点图
ggplot(inf_df, aes(x = subject_id, y = delta_beta)) +
  geom_point() +
  geom_segment(aes(xend = subject_id, yend = 0)) +
  labs(
    title = "Influence Diagnostics (Delta-Beta)",
    subtitle = "Change in 'Visit x High Graduation' coefficient when excluding each subject",
    y = "Change in Coefficient",
    x = "Subject ID"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) # 隐藏X轴标签以免拥挤
```

## 3.3 Robustness Check and Final Visualization

```{r}
#| label: gee-rq2-coef-comp
library(dplyr)
library(broom)
library(purrr)
library(tidyr)
models_rq2 <- list(
  Independence = gee_rq2_ind, 
  Exchangeable = gee_rq2_exch, 
  AR1 = gee_rq2_ar1
)

# 2. 提取系数并合并
coef_comp_rq2 <- map_dfr(models_rq2, tidy, .id = "model_name") %>%
  filter(term %in% c("(Intercept)", "visit", "age","sex","educ_catHigh","educ_catMedium","visit:educ_catHigh", "visit:educ_catMedium" 
                     )) %>%
  
  dplyr::select(model_name, term, estimate) %>%
  # 转换为宽格式以便对比
  pivot_wider(names_from = model_name, values_from = estimate)

# 3. 展示对比表格
knitr::kable(
  coef_comp_rq2, 
  digits = 3, 
  caption = "RQ2 : Robustness of MMSE Coefficient Estimates across Working Correlation Structures "
) %>%
  kable_styling(position = "center")
```

```{r}
#| label: gee-rq2-prediction-plot
library(ggplot2)

# 1. 选择最终模型 (基于经验相关性选择 Exchangeable)
final_model_rq2 <- gee_rq2_exch

# 2. 创建预测网格
# 固定协变量为均值/众数，只让 visit 和 group 变化
new_data_rq2 <- expand.grid(
  visit = oasis_gee$visit,
  group  = factor(c("Nondemented", "Converted", "Demented"), levels = levels(oasis_gee$group)),
  age    = mean(oasis_gee$age, na.rm = TRUE),
  educ_cat   = oasis_gee$educ_cat,
  sex    = factor("F", levels = levels(oasis_gee$sex)) # 以女性为例
)

# 3. 生成预测值
# 对于 Gaussian 模型，type="response" 返回的是预测的均值 (Predicted Mean MMSE)
preds_rq2 <- predict(final_model_rq2, newdata = new_data_rq2, type = "response")
new_data_rq2$pred_mmse <- as.numeric(preds_rq2)

# 4. 绘图
ggplot(new_data_rq2, aes(x = visit, y = pred_mmse, color = educ_cat, linetype = educ_cat)) +
  geom_line(size = 1.2) +
  scale_y_continuous(limits = c(0, 30)) + # MMSE 评分范围通常是 0-30
  labs(
    title = "Population-Average MMSE Trajectories Over Visit Time",
    subtitle = "Based on Gaussian GEE (Exchangeable Correlation)",
    y = "Predicted MMSE Score",
    x = "Visit",
    color = "educ_cat",
    linetype = "educ_cat"
  ) +
  theme_minimal()
```
